---
title: "EDA-Alex"
output: html_document
---

###Analyses by Alex
```{r Alex, echo=FALSE, warning=FALSE, message=FALSE, fig.width = 24, fig.height = 11, echo=FALSE}
#load libraries####
library(dplyr)
library(lme4)
library(lattice)
library(latticeExtra)
library(lsmeans)
library(MASS)
library(vegan)

##load functions####
#error
error.bar <- function(x, y, upper, lower=upper, length=0,...){
  if(length(x) != length(y) | length(y) !=length(lower) | length(lower) != length(upper))
    stop("vectors must be same length")
  arrows(x,y+upper, x, y-lower, angle=90, code=3, length=0, ...)
}

#se
se <- function(x) sd(x)/sqrt(length(x)) ## SE

#confidence intervals
conf.95 <- function(x) se(x)*2 #95% confidence interval bars

#rii function
rii <- function(x, j, var)
{
s1 <- subset(x, Microsite == "Shrub", select=var)
o1 <- subset(x, Microsite == "Open", select=var)
return1 <- (s1 - o1) / (s1+o1)
x1 <- x[seq(1, nrow(x), by = 2),]
return2 <- cbind(x1[j], return1)
return2[is.na(return2)] <- 0
return2
}

#data wrangling####
#primary data
data <- read.csv("data/shrub.contrast.survey.csv")
data.ord <- read.table("data/shrub.contrast.ordination.csv", header=T,sep=",")

#data for abundance and richness
data=data
data[is.na(data)] <- 0
#data for ordinations: 2015 & 2016 only
data.ord=data.ord
data.ord[is.na(data.ord)] <- 0

#generate riis
rii.dat <- rii(data, 1:8, c("abundance","richness")) #keep columns 1:8

##Parse by region and year
dat.2015 <- subset(data.ord, year==2015) ##subset data for 2015
pan.2015 <- subset(dat.2015, Region=="Panoche")
dat.2016 <- subset(data.ord, year==2016)
pan.2016 <- subset(dat.2016, Region=="Panoche")
mnp.2015 <- subset(dat.2015, Region=="Mojave")
mnp.2015 <- subset(mnp.2015, rowSums(mnp.2015[,11:41])>0) #remove empty plots for analysis
mnp.2016 <- subset(dat.2016, Region=="Mojave")
mnp.2016 <- subset(mnp.2016, rowSums(mnp.2016[,11:41])>0) #remove empty plots for analysis

#aggregate sites####
#site averages for each year
means <- aggregate(rii.dat, by=list(region=rii.dat$Region, site=rii.dat$Site,year=rii.dat$year), mean)
ses <- aggregate(rii.dat, by=list(region=rii.dat$Region, site=rii.dat$Site,year=rii.dat$year), conf.95)

#grand means for each year
grand.means <- aggregate(rii.dat, by=list(region=rii.dat$Region, site=rii.dat$Site,year=rii.dat$year), mean)
grand.ses <- aggregate(rii.dat, by=list(region=rii.dat$Region, site=rii.dat$Site,year=rii.dat$year), conf.95)

#Data viz Rii#####
##x-axis locations
xax <- seq(1,8, by=c(1))

#region names
Ephedra <- expression(italic(E.)*italic(californica))
Larrea <- expression(italic(L.)*italic(tridentata))

par(mfrow=c(2,4))
##abundance
##site 2013 abundance
par(mar=c(4.5,4.5,3,.5))
plot(xax, means[1:8,"abundance"], pch=19, ylim=c(-0.5,1), ylab="RII abundance", xlab="", cex=2.5, cex.axis=1.8, cex.lab=2, xlim=c(1,8), xaxt='n')
#axis(1, at=1:8, labels=c("south","plateau","north","313m","510m","730m","960m","1130m"), cex.axis=1.5)
axis(1, at=c(2,6), labels=c(Ephedra,Larrea), cex.axis=2, pos=-0.65)
axis(3, at=c(4.5), labels="2013", cex.axis=3)
error.bar(xax,means[1:8,"abundance"],ses[1:8,"abundance"], lwd=2.5)
abline(h=0, lty=2, lwd=2)
abline(v=3.5, lty=3, lwd=2)
## add grand means
rect(0,grand.means[2,"abundance"]-grand.ses[2,"abundance"],3.5,grand.means[2,"abundance"]+grand.ses[2,"abundance"],col="#80808030",border="transparent") #Panoche
rect(3.5,grand.means[1,"abundance"]-grand.ses[1,"abundance"],9,grand.means[1,"abundance"]+grand.ses[1,"abundance"],col="#80808030",border="transparent") #Mojave

##site 2014 abundance
par(mar=c(4.5,4.5,3,.5))
plot(xax, means[9:16,"abundance"], pch=19, ylim=c(-0.5,1), ylab="", xlab="", cex=2.5, cex.axis=1.8, cex.lab=2, xlim=c(1,8), xaxt='n')
#axis(1, at=1:8, labels=c("south","plateau","north","313m","510m","730m","960m","1130m"), cex.axis=1.5)
axis(1, at=c(2,6), labels=c(Ephedra,Larrea), cex.axis=2, pos=-0.65)
axis(3, at=c(4.5), labels="2014", cex.axis=3)
error.bar(xax,means[9:16,"abundance"],ses[9:16,"abundance"], lwd=2.5)
abline(h=0, lty=2, lwd=2)
abline(v=3.5, lty=3, lwd=2)
## add grand means
rect(0,grand.means[4,"abundance"]-grand.ses[4,"abundance"],3.5,grand.means[4,"abundance"]+grand.ses[4,"abundance"],col="#80808030",border="transparent") #Panoche
rect(3.5,grand.means[3,"abundance"]-grand.ses[3,"abundance"],9,grand.means[3,"abundance"]+grand.ses[3,"abundance"],col="#80808030",border="transparent") #Mojave

##site 2015 abundance
par(mar=c(4.5,4.5,3,.5))
plot(xax, means[17:24,"abundance"], pch=19, ylim=c(-0.5,1), ylab="", xlab="", cex=2.5, cex.axis=1.8, cex.lab=2, xlim=c(1,8), xaxt='n')
#axis(1, at=1:8, labels=c("south","plateau","north","313m","510m","730m","960m","1130m"), cex.axis=1.5)
axis(1, at=c(2,6), labels=c(Ephedra,Larrea), cex.axis=2, pos=-0.65)
axis(3, at=c(4.5), labels="2015", cex.axis=3)
error.bar(xax,means[17:24,"abundance"],ses[17:24,"abundance"], lwd=2.5)
abline(h=0, lty=2, lwd=2)
abline(v=3.5, lty=3, lwd=2)
## add grand means
rect(0,grand.means[6,"abundance"]-grand.ses[6,"abundance"],3.5,grand.means[6,"abundance"]+grand.ses[6,"abundance"],col="#80808030",border="transparent") #Panoche
rect(3.5,grand.means[5,"abundance"]-grand.ses[5,"abundance"],9,grand.means[5,"abundance"]+grand.ses[5,"abundance"],col="#80808030",border="transparent") #Mojave

##site 2016 abundance
par(mar=c(4.5,4.5,3,.5))
plot(xax, means[25:32,"abundance"], pch=19, ylim=c(-0.5,1), ylab="", xlab="", cex=2.5, cex.axis=1.8, cex.lab=2, xlim=c(1,8), xaxt='n')
#axis(1, at=1:8, labels=c("south","plateau","north","313m","510m","730m","960m","1130m"), cex.axis=1.5)
axis(1, at=c(2,6), labels=c(Ephedra,Larrea), cex.axis=2, pos=-0.65)
axis(3, at=c(4.5), labels="2016", cex.axis=3)
error.bar(xax,means[25:32,"abundance"],ses[25:32,"abundance"], lwd=2.5)
abline(h=0, lty=2, lwd=2)
abline(v=3.5, lty=3, lwd=2)
## add grand means
rect(0,grand.means[8,"abundance"]-grand.ses[8,"abundance"],3.5,grand.means[8,"abundance"]+grand.ses[8,"abundance"],col="#80808030",border="transparent") #Panoche
rect(3.5,grand.means[7,"abundance"]-grand.ses[7,"abundance"],9,grand.means[7,"abundance"]+grand.ses[7,"abundance"],col="#80808030",border="transparent") #Mojave

##richness 
##site 2013 richness
par(mar=c(4.5,4.5,.5,.5))
plot(xax, means[1:8,"richness"], pch=19, ylim=c(-0.5,1), ylab="RII richness", xlab="", cex=2.5, cex.axis=1.8, cex.lab=2, xlim=c(1,8), xaxt='n')
axis(1, at=1:8, labels=c("south","plateau","north","313m","510m","730m","960m","1130m"), cex.axis=1.5)
error.bar(xax,means[1:8,"richness"],ses[1:8,"richness"], lwd=2.5)
abline(h=0, lty=2, lwd=2)
abline(v=3.5, lty=3, lwd=2)
## add grand means
rect(0,grand.means[2,"richness"]-grand.ses[2,"richness"],3.5,grand.means[2,"richness"]+grand.ses[2,"richness"],col="#80808030",border="transparent") #Panoche
rect(3.5,grand.means[1,"richness"]-grand.ses[1,"richness"],9,grand.means[1,"richness"]+grand.ses[1,"richness"],col="#80808030",border="transparent") #Mojave

##site 2014 richness
par(mar=c(4.5,4.5,.5,.5))
plot(xax, means[9:16,"richness"], pch=19, ylim=c(-0.5,1), ylab="", xlab="", cex=2.5, cex.axis=1.8, cex.lab=2, xlim=c(1,8), xaxt='n')
axis(1, at=1:8, labels=c("south","plateau","north","313m","510m","730m","960m","1130m"), cex.axis=1.5)
error.bar(xax,means[9:16,"richness"],ses[9:16,"richness"], lwd=2.5)
abline(h=0, lty=2, lwd=2)
abline(v=3.5, lty=3, lwd=2)
## add grand means
rect(0,grand.means[4,"richness"]-grand.ses[4,"richness"],3.5,grand.means[4,"richness"]+grand.ses[4,"richness"],col="#80808030",border="transparent") #Panoche
rect(3.5,grand.means[3,"richness"]-grand.ses[3,"richness"],9,grand.means[3,"richness"]+grand.ses[3,"richness"],col="#80808030",border="transparent") #Mojave

##site 2015 richness
par(mar=c(4.5,4.5,.5,.5))
plot(xax, means[17:24,"richness"], pch=19, ylim=c(-0.5,1), ylab="", xlab="", cex=2.5, cex.axis=1.8, cex.lab=2, xlim=c(1,8), xaxt='n')
axis(1, at=1:8, labels=c("south","plateau","north","313m","510m","730m","960m","1130m"), cex.axis=1.5)
error.bar(xax,means[17:24,"richness"],ses[17:24,"richness"], lwd=2.5)
abline(h=0, lty=2, lwd=2)
abline(v=3.5, lty=3, lwd=2)
## add grand means
rect(0,grand.means[6,"richness"]-grand.ses[6,"richness"],3.5,grand.means[6,"richness"]+grand.ses[6,"richness"],col="#80808030",border="transparent") #Panoche
rect(3.5,grand.means[5,"richness"]-grand.ses[5,"richness"],9,grand.means[5,"richness"]+grand.ses[5,"richness"],col="#80808030",border="transparent") #Mojave

##site 2016 richness
par(mar=c(4.5,4.5,.5,.5))
plot(xax, means[25:32,"richness"], pch=19, ylim=c(-0.5,1), ylab="", xlab="", cex=2.5, cex.axis=1.8, cex.lab=2, xlim=c(1,8), xaxt='n')
axis(1, at=1:8, labels=c("south","plateau","north","313m","510m","730m","960m","1130m"), cex.axis=1.5)
error.bar(xax,means[25:32,"richness"],ses[25:32,"richness"], lwd=2.5)
abline(h=0, lty=2, lwd=2)
abline(v=3.5, lty=3, lwd=2)
## add grand means
rect(0,grand.means[8,"richness"]-grand.ses[8,"richness"],3.5,grand.means[8,"richness"]+grand.ses[8,"richness"],col="#80808030",border="transparent") #Panoche
rect(3.5,grand.means[7,"richness"]-grand.ses[7,"richness"],9,grand.means[7,"richness"]+grand.ses[7,"richness"],col="#80808030",border="transparent") #Mojave

# 2x2 plot####
par(mfrow=c(2,2))

## Hellinger transformation --> Legendre and Gallagher 2001
pan.2015.trans <-decostand(pan.2015[,11:40]+0.000001, method="hellinger") #select only species columns

##DCA with microsite as predictor
dca1 <- decorana(pan.2015.trans)

#Panoche 2015
par(mar=c(4.5,4.5,2.5,.5))
plot(dca1, xlim=c(-2.7,2.7), ylim=c(-2.7,2.7), dis="sp")
axis(3, 0, label="2015", cex.axis=1.8)

#create shrub-open labels for plot
ordi.lab <- rep(c("+","o"),45)# + - shrub, o - open

##display species
abu <- colSums(pan.2015.trans) #sum plots to prioritize species
ordilabel(scores(dca1)*2.5, labels=ordi.lab, dis="sites") #expand site locations on plot
ordilabel(dca1, col = "red", dis = "sp", priority = abu, cex=1.1) #plot species based on abundance

#Panoche 2016
pan.2016.trans <-decostand(pan.2016[,11:40]+0.000001, method="hellinger")
dca2 <- decorana(pan.2016.trans)

par(mar=c(4.5,4.5,2.5,2.5))
plot(dca2, xlim=c(-3,3), ylim=c(-2.7,2.7), dis="sp")
axis(3, 0, label="2016", cex.axis=1.8)
axis(4, 0, label="Panoche", cex.axis=1.8)

abu <- colSums(pan.2016.trans)
ordilabel(scores(dca2)*2.5, labels=ordi.lab, dis="sites") #expand site locations on plot
ordilabel(dca2, col = "red", dis = "sp", priority = abu, cex=1.1) #plot species based on abundance

#mojave 2015
mnp.2015.trans <-decostand(mnp.2015[,11:40], method="hellinger")
dca1 <- decorana(mnp.2015.trans)

par(mar=c(4.5,4.5,.5,.5))
plot(dca1, xlim=c(-3.5,3.5), ylim=c(-3.2,3.2), dis="sp")

##display species
abu <- colSums(mnp.2015.trans) #sum plots to prioritize species
ordilabel(scores(dca1)*1.2, labels=ordi.lab,dis="sites") #expand site locations on plot
ordilabel(dca1, col = "red", dis = "sp", priority = abu, cex=1.1) #plot species based on abundance

#mojave 2016
mnp.2016.trans <-decostand(mnp.2016[,11:40], method="hellinger")
dca2 <- decorana(mnp.2016.trans)

par(mar=c(4.5,4.5,.5,2.5))
plot(dca2, xlim=c(-3,3), ylim=c(-3,3), dis="sp")
axis(4, 0, label="Mojave", cex.axis=1.8)

##display species
abu <- colSums(mnp.2016.trans) #sum plots to prioritize species
ordilabel(scores(dca2)*1.2, labels=ordi.lab,dis="sites") #expand site locations on plot
ordilabel(dca2, col = "red", dis = "sp", priority = abu, cex=1.1) #plot species based on abundance

#Primary data####
m1 <- glm(abundance ~ year * Microsite * Region, data=data, family = poisson)
summary(m1)
m1$deviance/m1$df.residual

m1 <- glm.nb(abundance ~ year * Microsite * Region, data=data)
anova(m1, test="Chisq")

#Compare temporal and spatial component####
```{r fig.width=12, fig.height=7}
## spatial
rii.spatial <- rii(data, c("Site","Region"), c("abundance","richness"))

spatial <- rii.spatial %>% group_by(Site) %>% summarize(abd=mean(abundance),rich=mean(richness), abd.conf=conf.95(abundance), rich.conf=conf.95(richness))
spatial <- data.frame(spatial)

par(mfrow=c(1,2))
par(mar=c(4.5,4.5,.5,.5))

##plot abundance
plot(1:8,spatial[,"abd"], pch=19, cex=1.5, ylim=c(-0.4,0.8), xlab="Site", ylab="Rii Abundance", cex.axis=1.5, cex.lab=1.8)
error.bar(1:8,spatial[,"abd"],spatial[,"abd.conf"])
abline(h=0, lwd=2, lty=2)
##plot richness
plot(1:8,spatial[,"rich"], pch=19, cex=1.5, ylim=c(-0.4,0.8), xlab="Site", ylab="Rii Richness", cex.axis=1.5, cex.lab=1.8)
error.bar(1:8,spatial[,"rich"],spatial[,"rich.conf"])
abline(h=0, lwd=2, lty=2)

## Temporal
rii.temporal <- rii(data, c("year"), c("abundance","richness"))

temporal<- rii.temporal %>% group_by(year) %>% summarize(abd=mean(abundance),rich=mean(richness), abd.conf=conf.95(abundance), rich.conf=conf.95(richness))
temporal <- data.frame(temporal)

par(mfrow=c(1,2))
par(mar=c(4.5,4.5,.5,.5))

##plot abundance
plot(2013:2017,temporal[,"abd"], pch=19, cex=1.5, ylim=c(-0.4,0.8), xlab="year", ylab="Rii Abundance", cex.axis=1.5, cex.lab=1.8)
error.bar(2013:2017,temporal[,"abd"],temporal[,"abd.conf"])
abline(h=0, lwd=2, lty=2)
##plot richness
plot(2013:2017,temporal[,"rich"], pch=19, cex=1.5, ylim=c(-0.4,0.8), xlab="year", ylab="Rii Richness", cex.axis=1.5, cex.lab=1.8)
error.bar(2013:2017,temporal[,"rich"],temporal[,"rich.conf"])
abline(h=0, lwd=2, lty=2)
```

